---
title: The IMF discontinues Kenya's credit facility. Now what?
author: Edoardo
date: '2018-09-14'
slug: Currency devaluation
categories: []
tags: ["Macro", "Kenya"]
header:
  caption: ''
  image: ''
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(here)
library(tidyverse)
library(gganimate)
library(zoo)
```

Last week the IMF discontinued its $1.5 bil standby credit facility to Kenya. Although the Kenyan government [maintained](https://www.bloomberg.com/news/articles/2018-09-14/imf-says-kenya-failed-to-complete-second-review-of-program?srnd=economics-vp) that the economy is solid and it no longer needs the facility, it is more likely that the government would actually benefit from it, especially at a time when emerging markets are volatile, but did not agree with the IMF's condition to remove the interest rate cap. Since September 2016, Kenyan banks can lend at a maximum of 4% above the key lending rate.

*What is IMF's standby facility?*

The [IMF standby facility](https://www.imf.org/en/About/Factsheets/Sheets/2016/08/02/21/10/Standby-Credit-Facility) is a form of financial assistance provided by the IMF to low-income countries that have sound macroeconomic policies are at risk of short-term balance of payment needs. Other countries like [Rwanda](https://www.imf.org/en/News/Articles/2017/11/30/pr17457-rwanda-imf-executive-board-approves-extensions-of-the-standby-credit-facility-arrangement) and [Honduras](https://www.imf.org/en/News/Articles/2016/11/02/PR16482-Honduras-IMF-Completes-Third-Fourth-Reviews-Under-SBA-SCF) have accessed the fund as well in the last couple of years. Kenya agreed with the IMF on the standby facility in March 2016, and extended it in March 2018, but never actually used any of the funds.

*What is the balance of payments, and why does it matter for the standby facility?*

The balance of payment is essentially a metric that measures the monetary component of international trade. It accounts for all economic transactions between the residents of a country, e.g. Kenya, and the rest of the world. The metric includes imports and exports (balance of trade), as well as remittances, development aid and various types of financial flows.

The basic problem here is that Kenya has a negative balance of payments: it imports too much and exports too little. Since Kenya does not have enough exports money to finance its imports, it needs to obtain strong currency in other ways, such as development assistance, remittances and short-term capital inflows. The latter can be particular volatile, and can leave the country in need for support.

```{r import,echo=FALSE, message=FALSE, warning=FALSE}
qmfr_data <- readRDS(here("static", "data", "QMFR data Q22018.Rds"))

f <-qmfr_data %>% select(date, Imports=imports_total, Exports=exports_total, Balance=trade_balance) %>%
  filter(date>2005) %>%
  gather(key,value, (2:ncol(.))) %>%
  ggplot(aes(x=as.Date(date), y=value, col=key, group=key)) +
        geom_line() +
        labs(title="Kenya's trade (2005-2018", subtitle="Imports are growing at a faster rate than exports", y="KES bil", caption="Market prices", x="" )  +
        theme_minimal() +
  geom_text(aes(x = max(as.Date(date)-300), label = key), hjust = 0) + 
  transition_reveal(key, as.Date(date-1)) + 
  geom_segment(aes(xend = max(as.Date(date-1)), yend = value), linetype = 2, colour = 'grey') +
  coord_cartesian(clip = 'off') +
  geom_point(size = 2) +
  theme(legend.position="none")
  
animate(f)    
```

*The IMF standby facility is discontinued, now what?

Kenya never accessed the facility in the last two years. However, it was a form of insurance that gave confidence to investors and helped the government to attract foreign flows. Without this buffer, the flow of reserve currency (i.e. US Dollars) could be affected and a depreciation might ensue. The concern is fuelled by overall volatility in emerging markets registered in the last month.

The depreciation of the Kenyan Shilling however is not a new thing. A long-term analysis shows a somewhat constant downward trend in the last 20 years, with  improved stability in the last 3 years.  


```{r boh, echo=FALSE, message=FALSE, warning=FALSE}
qmfr_data %>% ggplot(aes(x=date, y=United.States.dollar)) +
        geom_line() +
        labs(title="KES vs USD", subtitle="The Kenyan Shilling has been decreasing in value, why?", y="KES per 1 USD" ) +
        scale_y_reverse() +
        theme_minimal()
        
```


But currency depreciation is not always a bad thing, especially when inflation levels are contained. Last year inflation was considerably below the average of the past 5 years. The question here is whether the conditions are conducive for exports to pick up speed

```{r inflation, echo=FALSE, message=FALSE, warning=FALSE}
#qmfr_data <- readRDS(here("static", "data", "QMFR data Q22018.Rds"))
qmfr_data %>% select(date, inflation_monthly) %>%
  filter(date>2005) %>%
  gather(key,value, (2:ncol(.))) %>%
  ggplot(aes(x=date, y=value)) +
        geom_line() +
        labs(title="Kenya's trade snapshot", subtitle="Imports are growing, exports are stagnant", y="%", caption="Market prices" )  +
        theme_minimal()
  
        
```

