---
title: Kenyan Shilling depreciation against the dollar, a long-term view
author: Edoardo
date: '2018-09-14'
slug: Currency devaluation
categories: []
tags: ["Macro", "Kenya"]
header:
  caption: ''
  image: ''
---

A long-term analysis of the Kenyan Shilling (KES) against the US Dollar (USD) leaves no doubt that the Kenyan Shilling has depreciated in the last 20 years, although it has stabilized in the last 3 years.  Kenya actively implemented several [economic reform measures (pdf)](https://aercafrica.org/wp-content/uploads/2018/07/rp109.pdf) since the 1990s inclusing a liberalization of interest rates (1990) and the exchange rate (1992/93).  

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(here)
library(tidyverse)
library(gganimate)
library(zoo)
```


```{r boh, echo=FALSE, message=FALSE, warning=FALSE}
qmfr_data <- readRDS(here("static", "data", "QMFR data Q22018.Rds"))
qmfr_data %>% ggplot(aes(x=date, y=United.States.dollar)) +
        geom_line() +
        labs(title="KES vs USD", subtitle="The Kenyan Shilling has been decreasing in value, why?", y="KES per 1 USD" ) +
        scale_y_reverse() +
        theme_minimal()
        
```

The reason for this decline is fairly simple: Kenya imports too much and exports too little. Since Kenya does not have enough exports money to finance its imports, it needs to obtain strong currency in other ways, such as short-term capital inflows, development assistance and remittance. There has been a steady growth of imports, while exports have grown at slow speed.

```{r import, echo=FALSE, message=FALSE, warning=FALSE}
#qmfr_data <- readRDS(here("static", "data", "QMFR data Q22018.Rds"))
qmfr_data %>% select(date, Imports=imports_total, Exports=exports_total, Balance=trade_balance) %>%
  filter(date>2005) %>%
  gather(key,value, (2:ncol(.))) %>%
  ggplot(aes(x=as.Date(date), y=value, col=key, group=key)) +
        geom_line() +
        labs(title="Kenya's trade (2005-2018", subtitle="Imports are growing at a faster rate than exports", y="KES bil", caption="Market prices", x="" )  +
        theme_minimal() +
  geom_text(aes(x = max(as.Date(date)-300), label = key), hjust = 0) + 
  transition_reveal(key, as.Date(date-1)) + 
  geom_segment(aes(xend = max(as.Date(date-1)), yend = value), linetype = 2, colour = 'grey') +
  coord_cartesian(clip = 'off') +
  geom_point(size = 2) +
  theme(legend.position="none")
        
```

But currency depreciation is not always a bad thing, especially when inflation levels are contained. Last year inflation was considerably below the average of the past 5 years. The question here is whether the conditions are conducive for exports to pick up speed

```{r inflation, echo=FALSE, message=FALSE, warning=FALSE}
#qmfr_data <- readRDS(here("static", "data", "QMFR data Q22018.Rds"))
qmfr_data %>% select(date, inflation_monthly) %>%
  filter(date>2005) %>%
  gather(key,value, (2:ncol(.))) %>%
  ggplot(aes(x=date, y=value)) +
        geom_line() +
        labs(title="Kenya's trade snapshot", subtitle="Imports are growing, exports are stagnant", y="%", caption="Market prices" )  +
        theme_minimal()
  
        
```

